---
title: "Supplemental Material for Lag Manuscript"
subtitle: "Lagged Predictions of Next Week Alcohol Use for Precision Mental Health Support"
author: "Kendra Wyant, Gaylen E. Fronk, Jiachen Yu, and John J. Curtin"
date: last-modified
number-sections: true
format: 
  html: 
    embed-resources: true
    toc: true 
    toc_depth: 5
editor_options: 
  chunk_output_type: console
---

This file contains the supplemental materials for *Lagged Predictions of Next Week Alcohol Use for Precision Mental Health Support*. It includes a transparency report and all supplemental figures and tables. Additional materials are made available on our study's OSF page ([https://osf.io/xta67/](https://osf.io/xta67/)).   

-----



```{r}
#| echo: false

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(source("https://github.com/jjcurtin/lab_support/blob/main/format_path.R?raw=true"))
suppressPackageStartupMessages(library(tidyposterior))
suppressPackageStartupMessages(library(kableExtra))
library(patchwork)


theme_set(theme_classic())

options(knitr.kable.NA = '')

path_models_lag <- format_path(str_c("studydata/risk/models/lag"))
path_processed <- format_path(str_c("studydata/risk/data_processed/lag"))
```


```{r}
#| echo: false


pp_tidy <- read_csv(here::here(path_models_lag, "pp_tidy.csv"), 
                                 show_col_types = FALSE) 

ci <- read_csv(here::here(path_models_lag, "test_metrics_all_pp_perf.csv"), 
                                 show_col_types = FALSE) |> 
  mutate(model = factor(model, levels = c("0 lag", "24 lag", "72 lag", "168 lag", "336 lag"),
                        labels = c("0 hours", "24 hours", "72 hours", "168 hours", "336 hours")))

pp_dem <- read_csv(here::here(path_models_lag, "pp_dem_all.csv"), 
                                 show_col_types = FALSE)

pp_dem_contrast <- read_csv(here::here(path_models_lag, "pp_dem_contrast_all.csv"), 
                                 show_col_types = FALSE)
```


### Figure S1: Full Posterior Distributions for auROC by Model

```{r}
#| label: fig-lag-posteriors
#| fig-cap: 
#|   - "The plots above depict the posterior probabilities for the areas under the receiver operating characteristic curves (auROCs) by model lag. Each row represents a model lag (0, 24, 72, 168, and 336 hours)). The solid vertical lines represent the median posterior probability. The horizontal lines represent the 95% CI."

  
pp_tidy |> 
  mutate(model = factor(model, levels = c("lag0", "lag24", "lag72", "lag168", "lag336"),
                        labels = c("0 hours", "24 hours", "72 hours", "168 hours", "336 hours"))) |>
  ggplot() + 
  geom_histogram(aes(x = posterior), fill = "light grey", color = "black", alpha = .4, 
                 bins = 30) +
  geom_segment(mapping = aes(y = 1200, yend = 1600, x = pp_median, xend = pp_median),
               data = ci) +
  geom_segment(mapping = aes(y = 1400, yend = 1400, x = pp_lower, xend = pp_upper),
                data = ci) +
  facet_wrap(~model, ncol = 1) +
  scale_y_continuous("Posterior Probability", breaks = c(0, 500, 1000, 1500)) +
  xlab("Area Under ROC Curve") +
  theme_classic() +
  theme(legend.position = "none")
```



### Table S1: Model performance by Demographic Group

```{r}
pp_sex <- pp_dem_contrast |> 
  filter(contrast == "female vs male") |> 
   mutate(ci = str_c("[", round(lower, 3), ", ", round(upper, 3), "]"),
         median = as.character(round(median, 3)),
         probability = as.character(round(probability, 3))) |> 
  select(lag, median, ci, probability) |> 
  rename(`Lag (hours)`= lag,
         Median = median,
         `Bayesian CI` = ci,
         Probability = probability)

pp_income <- pp_dem_contrast |> 
  filter(contrast == "below poverty vs above poverty") |> 
   mutate(ci = str_c("[", round(lower, 3), ", ", round(upper, 3), "]"),
         median = as.character(round(median, 3)),
         probability = as.character(round(probability, 3))) |> 
  select(lag, median, ci, probability) |> 
  rename(`Lag (hours)`= lag,
         `Median ` = median,
         `Bayesian CI ` = ci,
         `Probability ` = probability)

pp_race <- pp_dem_contrast |> 
  filter(contrast == "not white vs non-hispanic white") |> 
   mutate(ci = str_c("[", round(lower, 3), ", ", round(upper, 3), "]"),
         median = as.character(round(median, 3)),
         probability = as.character(round(probability, 3))) |> 
  select(lag, median, ci, probability) |> 
  rename(`Lag (hours)`= lag,
         `Median  ` = median,
         `Bayesian CI  ` = ci,
         `Probability   ` = probability)
```


```{r}
#| label: tbl-fairness
#| tbl-cap: "Model Performance by Demographic Group"

pp_sex |> 
  full_join(pp_income, by = "Lag (hours)") |> 
  full_join(pp_race, by = "Lag (hours)") |> 
  kbl() |> 
  add_header_above(c(" " = 1, "Race" = 3, "Sex" = 3, "Income" = 3))
```


